The URL for our Team GitHub repository is https://github.com/stevenH721/Yaowen-Hu.git
The URL to the Kernel and Kaggle dataset is https://www.kaggle.com/jboysen/ny-home-mortgage
title: "Interim Report I"
output: word_document
---
1. What is the data problem? What is the final managerial objective?



2. Describe the measurement types of each variable


3. Create a table summarizing the range or variation in each variable. Add statistics (mean,
median, standard deviation, etc.) as you deem necessary.
```{r}

```
4. How do you handle missing data in this dataset?
5. Provide histograms/density plots for key variables, such as action taken on loan, gender, ethnicity, etc.
```{r}
# Histogram for action taken on loan
homeMortgageStatus = homeMortgage %>% group_by(action_taken_name) %>%
summarise(CountOfActionTaken = n()) %>%
arrange(desc(CountOfActionTaken))

ggplot(homeMortgageStatus, aes(x = reorder(action_taken_name, CountOfActionTaken), y = CountOfActionTaken)) +
geom_bar(stat='identity',colour="white", fill ="yellow") +
geom_text(aes(x = action_taken_name, y=1,sep="",label=paste0(CountOfActionTaken)),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
labs(x = 'action_taken_name', y = 'Count Of Action Taken', title = 'Actions in Loans') +
coord_flip() + 
theme_bw()
```

```{r}
# Histogram for gender
homeMortgageGender = homeMortgage %>% group_by(applicant_sex_name) %>%
summarise(CountOfSex = n()) %>%
arrange(desc(CountOfSex))

ggplot(homeMortgageGender, aes(x = reorder(applicant_sex_name, CountOfSex), y = CountOfSex)) +
geom_bar(stat='identity',colour="white", fill ="yellow") +
geom_text(aes(x = applicant_sex_name, y=1,sep="",label=paste0(CountOfSex)),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
labs(x = 'Gender', y = 'Count', title = 'Histogram for Gender') +
coord_flip() + 
theme_bw()
```

```{r}
# Histogram for ethnicity
homeMortgageEthnicity = homeMortgage %>% group_by(applicant_ethnicity_name) %>%
summarise(CountOfEthnicity = n()) %>%
arrange(desc(CountOfEthnicity))

ggplot(homeMortgageEthnicity, aes(x = reorder(applicant_ethnicity_name, CountOfEthnicity), y = CountOfEthnicity)) +
geom_bar(stat='identity',colour="white", fill ="yellow") +
geom_text(aes(x = applicant_ethnicity_name, y=1,sep="",label=paste0(CountOfEthnicity)),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
labs(x = 'Ethinicity', y = 'Count', title = 'Histogram for Ethnicity') +
coord_flip() + 
theme_bw()
```

6. Create bivariate frequency distributions (tables or plots) for key variables
```{r}
# Bivariate frequency distribution for ethnicity
homeMortgageEthnicity2 = homeMortgage %>% group_by(action_taken_name,applicant_ethnicity_name) %>%
summarise(CountOfActionTaken = n()) 
homeMortgageEthnicity = homeMortgage %>% group_by(applicant_ethnicity_name) %>%
summarise(CountOfEthnicity = n()) 

homeMortgageEthnicity3 = inner_join(homeMortgageEthnicity,homeMortgageEthnicity2) %>%
mutate(percentage = (CountOfActionTaken/CountOfEthnicity) * 100 ) 

homeMortgageEthnicity3 %>%
  ggplot(aes(x = reorder(action_taken_name, percentage), 
                               y = percentage)) +
  geom_bar(stat='identity',colour="white", fill ="yellow") +
  facet_wrap(~ applicant_ethnicity_name) +
  geom_text(aes(x = action_taken_name, y = 1, label = paste0("(",round(percentage),"%)",sep="")),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
  labs(x = 'action_taken_name', y = '%age Count Of Action Taken', title = 'Actions in Loans') +
  coord_flip() + 
  theme_bw()

```

```{r}
# Bivariate frequency distribution for gender
homeMortgageGender2 = homeMortgage %>% group_by(action_taken_name,applicant_sex_name) %>%
summarise(CountOfActionTaken = n()) 
homeMortgageGender = homeMortgage %>% group_by(applicant_sex_name) %>%
summarise(CountOfSex = n()) 

homeMortgageGender3 = inner_join(homeMortgageGender,homeMortgageGender2) %>%
mutate(percentage = (CountOfActionTaken/CountOfSex) * 100 ) 

homeMortgageGender3 %>%
  ggplot(aes(x = reorder(action_taken_name, percentage), 
                               y = percentage)) +
  geom_bar(stat='identity',colour="white", fill ="yellow") +
  facet_wrap(~ applicant_sex_name) +
  geom_text(aes(x = action_taken_name, y = 1, label = paste0("(",round(percentage),"%)",sep="")),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
  labs(x = 'action_taken_name', y = '%age Count Of Action Taken', title = 'Actions in Loans') +
  coord_flip() + 
  theme_bw()
```

7. Discuss what the data patterns indicate, and what this could mean for the probability of
getting a loan approved.
```{r}

```




