---
output:
  word_document: default
  pdf_document: default
  html_document: default
---
The URL for our Team GitHub repository is https://github.com/stevenH721/Yaowen-Hu.git
The URL to the Kernel and Kaggle dataset is https://www.kaggle.com/jboysen/ny-home-mortgage
Title: Interim Report
Author: Yaowen Hu, Qingqing Cai, Yumei Xiong, Yu Shi
---

Summry
 


Variable Description and Statistics


Handling Missing Data
# Firstly, we need to decide whether the missing data is missing at random. For instance, applicants with lower income may not be willing to reveal their income level. 
## If the data is missing at random, we could simply remove the missing data sample since we have huge amount of observations. There are 2 types of deletion: listwise, 
# Listwise deletion (complete-case analysis) removes all data for an observation if there is one or more missing values. For example, in this dataset, most spaces in colomn V43 in blank, so we could delete this column for simplicity. Pairwise deletion analyses all cases in which the variables of interest are present and thus maximizes all data available by an analysis basis.
# If the data is missing not at random, we could replace missing data with substituted values. For instance, one of the choices is to replace the missing data with the mean of the data we have. Using this method, we could avoid circumstances such as default listwise deletion. 

```{r}
hmda = read.csv("ny_hmda_2015.csv")
hmda$Loan_Approval <- ifelse(hmda$action_taken==1|hmda$action_taken==2|hmda$action_taken==6,1,0)
```

```{r}
hmda$isMissingIncome <- ifelse(is.na(hmda$applicant_income_000s) == TRUE  ,1,0)
missingIncome <- xtabs(~isMissingIncome + Loan_Approval,hmda)
missingIncome
chisq.test(missingIncome)
```

```{r}
table(hmda$applicant_sex_name)
table(hmda$applicant_sex)
hmda$isMissingSex <- ifelse(hmda$applicant_sex == 3,1,0)
missingSex <-xtabs(~Loan_Approval + isMissingSex,hmda)
missingSex
chisq.test(missingSex)
```

```{r}
table(hmda$applicant_ethnicity_name)
table(hmda$applicant_ethnicity)
hmda$isMissingEthnicity <- ifelse(hmda$applicant_ethnicity == 3,1,0)
missingEthnicity <-xtabs(~isMissingEthnicity + Loan_Approval,hmda)
missingEthnicity
chisq.test(missingEthnicity)
```

```{r}
table(hmda$applicant_race_1)
hmda$isMissingRace <- ifelse(hmda$applicant_race_1 == 6,1,0)
missingRace<-xtabs(~isMissingRace + Loan_Approval,hmda)
missingRace
chisq.test(missingRace)
```

```{r}
hmda$applicant_sex_name[which(hmda$applicant_sex_name == "Information not provided by applicant in mail, Internet, or telephone application")] <- "Male"
hmda$applicant_ethnicity_name[which(hmda$applicant_ethnicity_name == "Information not provided by applicant in mail, Internet, or telephone application")] <- "Not Hispanic or Latino"
hmda$applicant_race_name_1[which(hmda$applicant_race_name_1 == "Information not provided by applicant in mail, Internet, or telephone application")] <- "White"
hmda$applicant_income_000s[is.na(hmda$applicant_income_000s)] <- 93.441091

homeMortgage<- hmda
```


Histogram Analysis
```{r}
install.packages("tidyverse")
install.packages("rpart")
install.packages("rpart.plot")
install.packages("caTools")
install.packages("caret")
install.packages("DT")

library(tidyverse)
library(rpart)
library(rpart.plot)
library(caTools)
library(caret)
library(DT)

rm(list=ls())
# Histogram for action taken on loan
homeMortgageStatus = homeMortgage %>% group_by(action_taken_name) %>%
summarise(CountOfActionTaken = n()) 
homeMortgageStatus %>%
ggplot(aes(x = action_taken_name, y = CountOfActionTaken)) +
geom_bar(stat='identity',colour="white", fill ="yellow") +
geom_text(aes(x = action_taken_name, y=1,label=paste0(CountOfActionTaken)),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
labs(x = 'action_taken_name', y = 'Count Of Action Taken', title = 'Histogram for action taken on loan') +
coord_flip() + 
theme_bw()
```

```{r}
# Histogram for gender
homeMortgageGender = homeMortgage %>% group_by(applicant_sex_name) %>%
summarise(CountOfActionTaken = n()) 
homeMortgageGender %>%
ggplot(aes(x = applicant_sex_name, y=CountOfActionTaken)) +
geom_bar(stat='identity',colour="white", fill ="yellow") +
geom_text(aes(y=1,label=paste0(CountOfActionTaken))) +
labs(x = 'Gender', y = 'Count Of Action Taken', title = 'Histogram for gender') +
coord_flip() + 
theme_bw()
```

```{r}
# Histogram for ethnicity
homeMortgageEthnicity = homeMortgage %>% group_by(applicant_ethnicity_name) %>%
summarise(CountOfActionTaken = n()) 
homeMortgageEthnicity %>%
ggplot(aes(x = applicant_ethnicity_name, y = CountOfActionTaken)) +
geom_bar(stat='identity',colour="white", fill ="yellow") +
geom_text(aes(x = applicant_ethnicity_name, y=100,label=paste0(CountOfActionTaken)),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
labs(x = 'Ethnicity', y = 'Count Of Action Taken', title = 'Histogram for ethnicity') +
coord_flip() + 
theme_bw()
```

```{r}
# Histogram for applicant's income
hist(homeMortgage$applicant_income_000s,main="Histogram of income",n=100,xlab="Applicant's income in thousands",ylab="Count",col="yellow",xlim=c(0,1500))
```


Bivariate relationship analysis

```{r}
# Bivariate frequency distribution for ethnicity
homeMortgageEthnicity = homeMortgage %>% 
group_by(applicant_ethnicity_name) %>%
summarise(CountOfEthnicity = n()) 

homeMortgageEthnicity2 = homeMortgage %>% 
group_by(action_taken_name,applicant_ethnicity_name) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageEthnicity3 = inner_join(homeMortgageEthnicity,homeMortgageEthnicity2) %>%
mutate(percentage = (CountOfActionTaken/CountOfEthnicity) * 100 ) 

homeMortgageEthnicity3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ applicant_ethnicity_name) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Ethnicity") +
  coord_flip() + 
  theme_bw()
```
From the graphs, we can see that applicants who are not Hispanic or Latino are more likely (57.7%) to have loan originated than those are Hispanic or Latino (48.8%). Also, applicants who are not Hispanic or Latino have a chance of 19.9% to have their application denied by financial institution, which is less than those who are Hispanic or Latino (25.2%).

```{r}
# Create income level
homeMortgage$Income.Level="0"
homeMortgage[homeMortgage$applicant_income_000s<=50,]$Income.Level="Low Income"
homeMortgage[homeMortgage$applicant_income_000s>50 & homeMortgage$applicant_income_000s<=150,]$Income.Level="Medium Income"
homeMortgage[homeMortgage$applicant_income_000s>150,]$Income.Level="High Income"

# Bivariate frequency distribution for income
homeMortgageIncome = homeMortgage %>% 
group_by(Income.Level) %>%
summarise(CountOfIncome = n()) 

homeMortgageIncome2 = homeMortgage %>% 
group_by(action_taken_name,Income.Level) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageIncome3 = inner_join(homeMortgageIncome,homeMortgageIncome2) %>%
mutate(percentage = (CountOfActionTaken/CountOfIncome) * 100 ) 

homeMortgageIncome3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ Income.Level) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Income") +
  coord_flip() + 
  theme_bw()
```
We divide applicants' income into three levels. Applicants whose income is lower than $50,000 are considered as low income people. Medium income applicants have income between $50,000 and $150,000, and high income applicants have income over $150,000.
From the graphs, we can see that high income applicants are most likely to have loan originated (59.4%), followed by medium income applicants (50.1%), and low income applicants are least likely to have loan originated (49.9%). When it comes to application denied by financial institution, the sequence is inverse, with probability of 14.1%, 16.5% and 29.4% for high, medium and low income applicants respectively. The reason is intuitively straightforward, that higher the income, the more guarantee of paying back loan, and the financial institution is more likely to originate loans with less worries of default.

```{r}
# Bivariate frequency distribution for race
homeMortgageRace = homeMortgage %>% 
group_by(applicant_race_name_1) %>%
summarise(CountOfRace = n()) 

homeMortgageRace2 = homeMortgage %>% 
group_by(action_taken_name,applicant_race_name_1) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageRace3 = inner_join(homeMortgageRace,homeMortgageRace2) %>%
mutate(percentage = (CountOfActionTaken/CountOfRace) * 100 ) 

homeMortgageRace3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ applicant_race_name_1) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Race") +
  coord_flip() + 
  theme_bw()
```
From the graphs, we can see that applicants who are White are most likely (58.5%) to have loan originated, while those who are American Indian or Alaska Native are least likely (39.1%). Also, applicants who are White or Aisan have the lowest chance of 19.4% to have their application denied by financial institution, while this percentage is 38.5% for those who are American Indian or Alaska Native.

```{r}
# Bivariate frequency distribution for gender
homeMortgageSex = homeMortgage %>% 
group_by(applicant_sex_name) %>%
summarise(CountOfSex = n()) 

homeMortgageSex2 = homeMortgage %>% 
group_by(action_taken_name,applicant_sex_name) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageSex3 = inner_join(homeMortgageSex,homeMortgageSex2) %>%
mutate(percentage = (CountOfActionTaken/CountOfSex) * 100 ) 

homeMortgageSex3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ applicant_sex_name) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Gender") +
  coord_flip() + 
  theme_bw()
```
From the graphs, we can see that gender has slight effects on actions in loans. Female applicants are slightly more likely (57.3%) to have loan originated than male applicants (57.1%). However, male applicants are less likely to have their application denied by financial institution, with the probability of 19.8%, which is less than female applicants (21.2%).

```{r}
# Bivariate frequency distribution for hoepa status
homeMortgageHoepa = homeMortgage %>% 
group_by(hoepa_status_name) %>%
summarise(CountOfHoepa = n()) 

homeMortgageHoepa2 = homeMortgage %>% 
group_by(action_taken_name,hoepa_status_name) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageHoepa3 = inner_join(homeMortgageHoepa,homeMortgageHoepa2) %>%
mutate(percentage = (CountOfActionTaken/CountOfHoepa) * 100 ) 

homeMortgageHoepa3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ hoepa_status_name) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Hoepa Status") +
  coord_flip() + 
  theme_bw()
```
From the graphs, we can see that whether a loan is subject to HOEPA has large effects on actions in loans. All of the HOEPA loans are originated, while only 51.9% of non-HOEPA loans are originated, with 18.1% of them denied by financial institution.

```{r}
# Bivariate frequency distribution for lien status
homeMortgageLien = homeMortgage %>% 
group_by(lien_status_name) %>%
summarise(CountOfLien = n()) 

homeMortgageLien2 = homeMortgage %>% 
group_by(action_taken_name,lien_status_name) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageLien3 = inner_join(homeMortgageLien,homeMortgageLien2) %>%
mutate(percentage = (CountOfActionTaken/CountOfLien) * 100 ) 

homeMortgageLien3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ lien_status_name) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Lien Status") +
  coord_flip() + 
  theme_bw()
```
From the graphs, we can see that loans secured by a first lien are most likely to be originated (61.1%), followed by loans secured by a subordinate lien (57.5%), and loans not secured by a lien are least likely to be originated (49.9%). When it comes to application denied by financial institution, the sequence is inverse, with probability of 19.1%, 30.3% and 44.4% for the above three types respectively. 

```{r}
# Bivariate frequency distribution for loan purpose
homeMortgageLoanPurpose = homeMortgage %>% 
group_by(loan_purpose_name) %>%
summarise(CountOfLoanPurpose = n()) 

homeMortgageLoanPurpose2 = homeMortgage %>% 
group_by(action_taken_name,loan_purpose_name) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageLoanPurpose3 = inner_join(homeMortgageLoanPurpose,homeMortgageLoanPurpose2) %>%
mutate(percentage = (CountOfActionTaken/CountOfLoanPurpose) * 100 ) 

homeMortgageLoanPurpose3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ loan_purpose_name) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Loan Purpose") +
  coord_flip() + 
  theme_bw()
```
From the graphs, we can see that loans used for home purchase are most likely to be originated (58.0%), followed by loans used for home improvement (51.6%), and loans used for refinancing are least likely to be originated (44.3%). When it comes to application denied by financial institution, loans used for home purchase are least likely to be denied (10.1%), followed by purpose of refinancing (23.4%) and home improvement (36.4%).

```{r}
# Create loan amount level
homeMortgage$LoanAmount.Level="0"
homeMortgage[homeMortgage$loan_amount_000s<=100,]$LoanAmount.Level="Low Loan Amount"
homeMortgage[homeMortgage$loan_amount_000s>100 & homeMortgage$loan_amount_000s<=350,]$LoanAmount.Level="Medium Loan Amount"
homeMortgage[homeMortgage$loan_amount_000s>350,]$LoanAmount.Level="High Loan Amount"

# Bivariate frequency distribution for loan amount
homeMortgageLoanAmount = homeMortgage %>% 
group_by(LoanAmount.Level) %>%
summarise(CountOfLoanAmount = n()) 

homeMortgageLoanAmount2 = homeMortgage %>% 
group_by(action_taken_name,LoanAmount.Level) %>%
summarise(CountOfActionTaken = n()) 

homeMortgageLoanAmount3 = inner_join(homeMortgageLoanAmount,homeMortgageLoanAmount2) %>%
mutate(percentage = (CountOfActionTaken/CountOfLoanAmount) * 100 ) 

homeMortgageLoanAmount3 %>%
  ggplot(aes(x = action_taken_name, y = percentage)) +
  facet_wrap(~ LoanAmount.Level) +
  geom_bar(stat="identity",fill="lightblue") +
  geom_text(aes(x = action_taken_name, y = 1, label = ifelse(floor(round(percentage,1))==round(percentage,1),paste0(round(percentage,1),".0%"),paste0(round(percentage,1),"%"))),
            hjust=0,size = 4.5, colour = "black") +
  labs(x = "action_taken_name", y = "Percentage Count Of Action Taken", title = "Actions in Loans by Loan Amount") +
  coord_flip() + 
  theme_bw()
```
We divide loan amount into three levels. Loan amount which is lower than $100,000 are considered as low loan amount. Medium loan amount has a range between $100,000 and $350,000, and high loan amount is over $350,000.
From the graphs, we can see that high amount loans are most likely to be originated (52.4%), followed by medium amount loans (51.7%), and low amount loans are least likely to be originated (51.5%). When it comes to application denied by financial institution, the probability of being denied is 14.9%, 26.3% and 15.8% for the three types of loans respectively. The results are counterintuitive, and the reason may be that the loan amount that people apply for is correlated with other variables, such as loan purpose and applicant's income. The graphs here only reflect the relationship between actions in loan and loan amount, without taking into control other variables. 

Model Regression
Original model
Adjusted model
```{r}
install.packages("rsq")
model = glm(homeMortgage$Loan_Approval ~ homeMortgage$applicant_income_000s + homeMortgage$loan_amount_000s + factor(homeMortgage$applicant_ethnicity) + factor(homeMortgage$applicant_race_1) + factor(homeMortgage$applicant_sex) + factor(homeMortgage$lien_status) + factor(homeMortgage$loan_purpose) + homeMortgage$applicant_income_000s*factor(homeMortgage$applicant_ethnicity) +  homeMortgage$loan_amount_000s*factor(homeMortgage$loan_purpose),family = binomial(link = "logit"),data = homeMortgage)
summary(model)

#Calculate adjusted R squared
rsq::rsq(model,adj=TRUE)
```

```{r}
# Divide the dataset into training data and testing data, with 80% and 20% of data respectively
randOrder=order(runif(nrow(homeMortgage)))
trainingData=subset(homeMortgage,randOrder<0.8*nrow(homeMortgage))
testData=subset(homeMortgage,randOrder>=0.8*nrow(homeMortgage))

# Calculate MSE using test data
mean((testData$Loan_Approval-predict(model,testData))^2)
```
This is the final model we choose. We include the interaction between applicant's income and ethnicity, and that between loan amount and loan purpose, because we think applicant's income is correlated with ethnicity, and loan amount is correlated with loan purpose. We also tried to include other interactions, but most of the coefficients are not significant. Also, we first take into account HOEPA status because we originally think it's important. But after we run the regression, we find the coefficient is not siginificant, so we choose to exclude this variable. 
MSE is 46.58, which is the lowest among all the models that we think are economically reasonable, suggesting that the model predicts well. Adjusted R squared is 13.98%, which is relatively high (although it's not the highest, because we don't include all the variables and interactions).
From the model, we can see that applicant's income, loan amount, ethnicity, race, gender, lien status and loan purpose all have significant effects on whether a loan can be approved. 
-applicant_income_000s
Applicant's income is positively related to loan approval. Keeping other variables constant, when applicant's income increases by $1,000, the log of odds ratio of loan approval will increase by 0.00035.
-loan_amount_000s
Loan amount is nagetively related to loan approval. Keeping other variables constant, when loan amount increases by $1,000, the log of odds ratio of loan approval will decrease by 0.00003.
-applicant_ethnicity
Applicant ethnicity has a significant effect on loan approval, but the effect is minimal.
-applicant_race_1
Applicant race has a significant effect on loan approval. For example, keeping other variables constant, compared with American Indian or Alaska Native, Black or African American will have the log of odds ratio of loan approval that is 0.003 higher.
-applicant_sex
The effect of gender on loan approval is not significant, as the coefficient of factor(applicant_sex)2 is not significant.
-lien_status
The coefficients of factor(lien_status)2 and factor(lien_status)4 are not significant, and the coefficient of factor(lien_status)3 is significant, but the effect is minimal.
-loan_purpose
Loan purpose has a significant effect on loan approval, but the effect is minimal.
-applicant_income_000s*applicant_ethnicity
For example, compared with Hispanic or Latino applicants, applicants who are not Hispanic or Latino will make the slope (the effect of applicant's income on the log of odds ratio of loan approval) increase by 0.0005.
-loan_amount_000s*loan_purpose
For example, compared with home purchase, loans used for home improvement will make the slope (the effect of loan amount on the log of odds ratio of loan approval) increase by 0.000009, which is quite minimal.


Conclusion and Recommendation

